Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-21 Paper/Paper.tex
table and references updated

\maketitle
\begin{abstract}
Activity recognition solely with an accelerometer is a common investigated research topic and enables the detection of basic activities like sit, walk or stand. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition.  We present a sensor set-up that consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. open door, prepare dinner, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{ambient assisted living, capacitive proximity sensors, activity recognition, user context}
\end{abstract}

@@

This paper is structured as follows: We give a short insight in related work in chapter 2, followed by describing the hardware set-up used for this study in chapter 3. The experiment and evaluation will be described in chapter 4, showing how data has been obtained and how activity recognition is being improved. This work is concluded in chapter 5 with an outlook to future work.

\section{Related Work}

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{While entering the door, the wristband approaches the door knob twice, while opening and closing. Regarding the capacitive proximity sensor, this situation can be observed at the beginning and at the end of the activity.}
	\label{fig:tobias_1}
\end{figure}

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_2.pdf}
	\caption{Example of a ``sitting'' activity in which user changed his position very frequently. Throughout the whole activity the values of the proximity sensor stay constant due to the close proximity to the couch's fabric.}
	\label{fig:eugen_2}
\end{figure}

@@

\begin{figure}[htpb]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a toast. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows reoccurring movements.}
	\label{fig:eugen_5}
\end{figure}

@@

\begin{figure}[htpb]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An eating person who took 5 bites from the bread. After each bite, the hand is placed on the table.}
	\label{fig:eugen_6}
\end{figure}

@@

\begin{figure}[htpb]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_7.pdf}
	\caption{An example of the ``drinking'' activity of a test person. The participant first pours some water into the glass and then takes three drams of water. After each dram, he returns his arm to the table which can be observed in the characteristic patterns of the proximity sensor.}
	\label{fig:eugen_7}
\end{figure}

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/marko_8.pdf}
	\caption{An examplary instance of the class ``walking''. The acceleration sensor and the proximity sensor show periodic reoccuring patterns that are related to the pendulum-like arm movement and the proximity to the person's body during those movements.}
	\label{fig:marko_8}
\end{figure}

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant. One draw conclusions about the coverage of the arm with either couchons, blankets or the proximity to the mattress.}
	\label{fig:eugen_9}
\end{figure}

@@

\begin{table}[htbp]
	\centering	\begin{tabular}{p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{2.2cm}}
a & b & c & d & e & f & g & h & i & a & b & c & d & e & f & g & h & i & $\leftarrow$ classified as\\
\hline
10 & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 		& 9 & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & a = open door\\
\hline
2 & 259 & 4 & 2 & 4 & 2 & 4 & 0 & 0 	& 1 & 248 & 5 & 6 & 4 & 2 & 5 & 0 & 0 & b = sit\\
\hline
0 & 5 & 213 & 2 & 2 & 1 & 8 & 0 & 4 	& 0 & 3 & 212 & 3 & 2 & 1 & 6 & 0 & 8 & c = lying\\
\hline
3 & 5 & 2 & 52 & 9 & 3 & 1 & 1 & 1 		& 0 & 8 & 2 & 44 & 7 & 8 & 1 & 1 & 3 & d = get things\\
\hline
1 & 9 & 3 & 13 & 46 & 6 & 3 & 0 & 0 	& 1 & 7 & 1 & 11 & 37 & 9 & 2 & 0 & 10 & e = make bread\\
\hline
0 & 5 & 1 & 1 & 5 & 62 & 5 & 0 & 1 		& 0 & 3 & 2 & 3 & 8 & 58 & 5 & 0 & 1 & f = eat\\
\hline
0 & 8 & 4 & 2 & 1 & 5 & 68 & 0 & 0 		& 0 & 8 & 3 & 0 & 6 & 4 & 67 & 0 & 1 & g = drink\\
\hline
1 & 0 & 0 & 1 & 0 & 0 & 0 & 12 & 0 		& 1 & 0 & 0 & 1 & 0 & 0 & 0 & 13 & 0 & h = walk\\
\hline
1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & 325 	& 0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & 312 & i = sleep\\
		\end{tabular}
	\caption{Comparison of two Confusion matrices for an exemplary user. The left part includes the capacitive proximity sensor, the right part represents classification results without the sensor.}
	\label{tab:ConfusionMatrixForAnExemplaryUser}
\end{table}

Table \ref{tab:confmatrix}

\section{Conclusion and Outlook}

>>>>

\maketitle
\begin{abstract}
Activity recognition solely with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition. This work presents a sensor prototype which consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{ambient assisted living, capacitive proximity sensors, activity recognition, user context}
\end{abstract}

@@

% This paper is structured as follows: We give a short insight in related work in chapter 2, followed by describing the hardware set-up used for this study in chapter 3. The experiment and evaluation will be described in chapter 4, showing how data has been obtained and how activity recognition is being improved. This work is concluded in chapter 5 with an outlook to future work.

The remainder of this paper is  is structured as follows: First, Section \ref{sect:related} is dedicated to related work work in activity recognition, in particular considering approaches with multiple sensor modalities. Section \ref{sect:hardware} will present the hardware based on which a wrist-worn sensor prototype was built, fusing together an inertial data logger and a capacitive sensor integrated in a wristband. The experimental setup with the scripted scenario of daily activities and the activity recognition evaluation results, showing the performance boost of the capacitive sensor unit, are given in Section \ref{sect:experiment}. The paper is wrapped up with a conclusions section enumerating the findings of the paper, as well as pointing out future research potential.

\section{Related Work}
\label{sect:related}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The works of \cite{Fishkin_2005} and \cite{patterson2005fga} show that detecting objects the user touches or uses can be very helpful for activity recognition. By using RFID readers that are embedded in gloves or bracelets at the wrist and RFID tags attached to various objects of interest, one can detect the object grasped and used by the user, thus aiding the activity recognition in various application scenarios, such as activities of daily living \cite{phealth:maja} and \cite{Philipose_2004}, activity tracking in car manufacturing \cite{Stiefmeier08}, or household and gardening activities \cite{berlin_laerhoven_tei_2010}.

Our approach to enhance the inertial data from a wearable sensor is quite similar to the RFID scenarios just mentioned, as it also relies on a single wearable sensor and thus is aiming at an unobtrusive deployment. The main difference lies in the fact that we do not need accurate detection of tagged objects, but are only interested in the proximity to various objects which, as we expect, could boost activity recognition.

% While the RFID gloves or bracelets might not have reached user acceptance, and deploying tags in the user's home might appear cumbersome and still quite intrusive, our approach considers a single wrist-worn sensor unit consisting of an accelerometer and proximity sensor.
% Berlin et al. \cite{berlin_laerhoven_tei_2010} have also fused and benchmarked a wrist-worn accelerometer sensor node with an RFID bracelet to be able to detect objects a human is interacting with, aiming both at household as well as gardening activities.

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{While entering the door, the wristband approaches the door knob twice, while opening and closing. Regarding the capacitive proximity sensor, this situation can be observed at the beginning and at the end of the activity.}
	\label{fig:tobias_1}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_2.pdf}
	\caption{Example of a ``sitting'' activity in which user changed his position very frequently. Throughout the whole activity the values of the proximity sensor stay constant due to the close proximity to the couch's fabric.}
	\label{fig:eugen_2}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a toast. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows reoccurring movements.}
	\label{fig:eugen_5}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An eating person who took 5 bites from the bread. After each bite, the hand is placed on the table.}
	\label{fig:eugen_6}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_7.pdf}
	\caption{An example of the ``drinking'' activity of a test person. The participant first pours some water into the glass and then takes three drams of water. After each dram, he returns his arm to the table which can be observed in the characteristic patterns of the proximity sensor.}
	\label{fig:eugen_7}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/marko_8.pdf}
	\caption{An examplary instance of the class ``walking''. The acceleration sensor and the proximity sensor show periodic reoccuring patterns that are related to the pendulum-like arm movement and the proximity to the person's body during those movements.}
	\label{fig:marko_8}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant. One draw conclusions about the coverage of the arm with either couchons, blankets or the proximity to the mattress.}
	\label{fig:eugen_9}
\end{figure}

\subsection{Evaluation Results}
\label{sect:evaluation}

@@

\begin{table}
	\caption{Comparison of two Confusion matrices for an exemplary user. The left part includes the capacitive proximity sensor, the right part represents classification results without the sensor.}

	\setlength{\tabcolsep}{2.4pt}
	\centering

	\begin{scriptsize}
	\begin{tabular}%{p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{2.2cm}}
	{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
		\multicolumn{9}{c}{with proximity} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{without proximity} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
		\textbf{10} & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 	&& \textbf{9} & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 		&& a = open door	\\ 
		2 & \textbf{259} & 4 & 2 & 4 & 2 & 4 & 0 & 0 	&& 1 & \textbf{248} & 5 & 6 & 4 & 2 & 5 & 0 & 0 	&& b = sitting		\\ 
		0 & 5 & \textbf{213} & 2 & 2 & 1 & 8 & 0 & 4 	&& 0 & 3 & \textbf{212} & 3 & 2 & 1 & 6 & 0 & 8 	&& c = lying		\\ 
		3 & 5 & 2 & \textbf{52} & 9 & 3 & 1 & 1 & 1 	&& 0 & 8 & 2 & \textbf{44} & 7 & 8 & 1 & 1 & 3 		&& d = get things	\\ 
		1 & 9 & 3 & 13 & \textbf{46} & 6 & 3 & 0 & 0 	&& 1 & 7 & 1 & 11 & \textbf{37} & 9 & 2 & 0 & 10 	&& e = make bread	\\ 
		0 & 5 & 1 & 1 & 5 & \textbf{62} & 5 & 0 & 1 	&& 0 & 3 & 2 & 3 & 8 & \textbf{58} & 5 & 0 & 1 		&& f = eating		\\ 
		0 & 8 & 4 & 2 & 1 & 5 & \textbf{68} & 0 & 0 	&& 0 & 8 & 3 & 0 & 6 & 4 & \textbf{67} & 0 & 1 		&& g = drinking		\\
		1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{12} & 0 	&& 1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{13} & 0 		&& h = walking		\\
		1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & \textbf{325} 	&& 0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & \textbf{312} 	&& i = sleeping		\\ 
		\cline{1-9} \cline{11-19} 
	\end{tabular}
	\end{scriptsize}

	\label{tab:ConfusionMatrixForAnExemplaryUser}
\end{table}

Table \ref{tab:ConfusionMatrixForAnExemplaryUser}

\section{Conclusion and Outlook}
\label{sect:conclusions}
